---
title: "Introduction to the `cci` package"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Introduction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
set.seed(1)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette introduces the `cci` package, and how to use it to calculate the
Charlson Comorbidity Index (CCI) based on diagnosis data using the International 
Statistical Classification of Diseases and Related Health Problems (ICD).

The package is inspired by the excellent `comorbidiy` [package](https://ellessenne.github.io/comorbidity/), but is focused on 
facilitating calculation of the CCI when working with data from the Danish
National Patient Registry that uses ICD-8 and ICD-10 diagnosis codes.

This vignette uses the following packages:

```{r}
library(cci)
library(data.table)
```


## International Statistical Classification of Diseases and Related Health Problems (ICD)

Different versions of ICD codes exists. In the Danish National Patient Registry
(DNPR) diagnoses are coded using ICD-8 prior to 1994, and as ICD-10 afterwards.
Because of this, the package focuses on these versions of the ICD codes, but it
is possible to modify/overwrite the default CCI definition so that eg ICD-9 
codes can be included in the definition. See the Codelists section.


## Sample ICD codes using `sample_codes()`

The package includes a helper function `sample_codes()`, that can be used to
simulate diagnosis data by sampling from lists of ICD codes included with
the package. 

```{r}
sample_codes(10L)
```

By default, the function samples both ICD-8 and ICD-10. If only want to sample
other combinations of ICD codes, this can be done using the `code_sources`
argument. 

```{r}
sample_codes(10L, code_sources = "icd10_cm_2020")
```

Currently, only ICD-8 and ICD-10 can be sampled. Finally, we can control the 
maximum length of the code that is sampled, so that if a very specific code is 
sampled, it is truncated:

```{r}
sample_codes(10L, max_code_length = 2)
```


## Calculate the CCI

The main function of the `cci` package is `calculate_cci()`. It takes as input
a data.frame with diagnosis data, and calculates the CCI for unique values of
the id variable that identifies each person.

```{r}
diag <- data.frame(
    id = rep(1:5, each = 20),
    code = sample_codes(100)
  )

head(diag)

calculate_cci(diag)
```

The above function calls assume that the id and diagnosis variable names are 
called "id" and "code". If this is not the case, the correct variable names 
can be specified using the `id`and `code` arguments.


## Corrections

## Handling SKS codes in the Danish National Patient Registry

As mentioned, this package is focused on facilitating use of data from the DNPR.
Diagnosis codes in the DNPR are so-called SKS-codes, which means in practice
that all ICD-10 codes have a "D" prefix. The `calculate_cci` function 
automatically converts SKS ICD-10 codes to real ICD-10 codes. If the diagnosis
data does not contain SKS ICD-10 codes, consider setting `sks_codes = FALSE`
(informal testing suggests this does not make a difference in practice).


## Codelists

What ICD codes to include/exclude in the definition of the Charlson Comorbidity
Index is not at all obvious. First of all, it obviously depends on what version(s)
of ICD codes are used in the diagnosis data, but it can also depend on how the codes are
used in practice. For example, the use of ICD-10 codes in Denmark differ from
the ICD-10(-CM) codes. Exactly how is not clear to the author of this package
(at least as of writing this), but from conversations with epidemiologists "Z"
codes are apparently used differently, and some "augmentations" have been made
to some ICD-10 codes by adding suffixes like "M" to cancer codes to indicate
metastasis. The point is, that there is likely not one "correct" definition of
codes to use when calculating the CCI, and the user needs to be very aware of 
what codes are used, and if those are appropriate for data at hand.

The `calculate_cci` function has a `codelist` argument, that can be used to
overwrite the default definition. The default codelist `cci_codelist_icd8_icd10`
uses both ICD-8 and ICD-10 codes, and is a typical definition used in 
observational studies using diagnosis data from the Danish National Patient Registry.
The object is included in the package and can easily be modified/overwitten as needed.

```{r}
diag <- data.frame(
  id = rep(1:5, each = 200),
  code = sample_codes(1000)
)

# Use default codelist
cci_default <- calculate_cci(diag)
# Use codelist only using ICD-10 codes
cci_only_icd10 <- calculate_cci(diag, codelist = cci::cci_codelist_icd10)

# The codelist object is just a nested list, with a list element for each 
# disease group in the CCI definition
cci_codelist_icd8_icd10[1]
# So we could make a modified list where we change the MI definition like so
modified_list <- cci_codelist_icd8_icd10
modified_list[[1]][["codes"]] <- c("Some set of codes")
cci_mod1 <- calculate_cci(diag, codelist = modified_list)

```


## Example

We will end this vignette with showing an example of how the package could be
used in a real study. First, lets simulate a study population with an index date
and diagnosis data on this population.

```{r}
pop <- data.table(
  id = 1:10,
  index_date = sample.int(100, 10, replace = TRUE)
)

diag <- data.table(
  id = rep(1:10, each = 50),
  code = sample_codes(500),
  code_date = sample.int(100, 500, replace = TRUE)
)


```

Typically, we would be interested in calculating the CCI on the index date, 
using all diagnosis data prior to the index date, but not on or after that date.
Finally, we would like merge the calculated CCI to the study population. This 
could be achieved like this:

```{r}
# Merge diagnosis data to population, restrict to diagnoses before the
# index date, and calculate the CCI.
cci <- merge(pop, diag, by = "id")[code_date < index_date, ] |>
  calculate_cci()
# Merge the result to the study population, and take into account that some
# patients might not have any relevant diagnosis data.
cci <- merge(pop, cci, by = "id", all.x = TRUE)[, cci := ifelse(is.na(cci), 0, cci)]
```










